\documentclass[11pt,a4paper]{article}

% ============================================================
% Predicting Albanian Student Low-Proficiency Risk from OECD PISA (2009-2022)
% Single-column academic report layout: title page, table of contents,
% running headers, numbered sections. (Previously an acmart manuscript.)
% ============================================================

\usepackage[margin=1in]{geometry}
\usepackage{booktabs}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{amsmath,amssymb}
\usepackage{array}
\usepackage{xcolor}
\usepackage{caption}
\usepackage[round,authoryear]{natbib}
\usepackage{fancyhdr}
\usepackage{titlesec}
\usepackage{setspace}
\usepackage{lineno}
\usepackage[hidelinks]{hyperref}

\graphicspath{{figures/}}

% Colored inline statistic highlights (crisis = red, recovery/proficient = green).
\definecolor{riskred}{RGB}{170,32,32}
\definecolor{okgreen}{RGB}{20,110,60}
\newcommand{\athigh}[1]{\textcolor{riskred}{#1}}
\newcommand{\atlow}[1]{\textcolor{okgreen}{#1}}

% Running headers on every body page.
\pagestyle{fancy}
\fancyhf{}
\fancyhead[L]{\small\itshape Predicting Albanian Low-Proficiency Risk}
\fancyhead[R]{\small\itshape PISA 2009--2022}
\fancyfoot[C]{\thepage}
\renewcommand{\headrulewidth}{0.4pt}

% Section heading style: "1. Title" with a rule underneath, report-style.
\titleformat{\section}
  {\large\bfseries}{\thesection.}{0.6em}{}[\vspace{2pt}\hrule]
\titleformat{\subsection}
  {\normalsize\bfseries}{\thesubsection}{0.6em}{}
\titlespacing*{\section}{0pt}{1.4em}{0.7em}

\captionsetup{font=small,labelfont=bf}
\setlength{\parskip}{0.4em}
\setlength{\parindent}{0pt}

% Journal submission formatting: double line spacing + continuous line numbers.
\doublespacing

% Table of contents lists sections only, so it fits on one page.
\setcounter{tocdepth}{1}

\begin{document}

% =========================== TITLE PAGE ===========================
\begin{titlepage}
\centering
\vspace*{0.9cm}

{\LARGE\bfseries A Structural Break, Not a Weaker Cohort\par}
\vspace{0.5cm}
{\large Explaining Albania's 2022 PISA Collapse with\\Survey-Weighted Machine Learning\par}
\vspace{0.8cm}
{\large PISA 2009--2022 Data Science Project\par}
\vspace{0.9cm}

{\large Xhoi Ikonomi\par}
\vspace{0.35cm}
{MSc Artificial Intelligence (Data Science profile)\\[2pt]
Metropolitan University of Tirana\\[2pt]
\texttt{ikonomixhoja@gmail.com}\par}

\vspace{0.8cm}
\rule{0.42\linewidth}{0.4pt}\par
\vspace{0.4cm}
{\bfseries Abstract\par}
\vspace{0.25cm}
\begin{minipage}{0.86\linewidth}
\small
\textbf{Background.}
Albania's share of 15-year-olds below PISA mathematics Level~2 rose from
\atlow{42\%} in 2018 to \athigh{75\%} in 2022, the sharpest single-cycle regression in
the assessment. We ask whether machine learning on background-questionnaire data can
identify at-risk students, why they are at risk, and whether the 2022 collapse is merely
a lower-scoring cohort or a change in the risk structure itself.

\textbf{Methods.}
Using all five Albanian PISA cycles (2009--2022; 27{,}042 students) and eight comparison
countries (323{,}121 students), we apply the survey methodology PISA requires: final
student weights on every estimate, plausible values combined by Rubin's rules, and
balanced repeated replication for design-based standard errors. We compare
survey-weighted classifiers, add school socioeconomic composition features, cross-check a
multilevel model and a direct school-questionnaire linkage, and audit calibration,
fairness, coverage and a difference-in-differences.

\textbf{Results.}
A domain classifier separates 2018 from 2022 students with AUC~0.98 on background
features alone, so the collapse is a structural covariate shift, not just lower scores.
Individual background variables hit a ceiling (cross-validated AUC~0.73) that tuning does
not move; adding school socioeconomic composition lifts every model to $\sim$0.78 and
makes gradient boosters beat logistic regression, exposing a \emph{feature} ceiling. The
new $\sim$0.78 plateau is then a genuine \emph{data} ceiling: a multilevel model (school
ICC~$\approx$~0.31) and a school-questionnaire linkage (no gain) converge on it, it
recurs across all nine systems with its size tracking each system's between-school ICC
($r\approx0.80$), and richer within-test process data reaches $\sim$0.86 only through a
signal endogenous to the score. The crisis is broad-based: Albania has the flattest
socioeconomic gradient of the nine countries, shares its risk drivers with peers, and a
difference-in-differences with a failed placebo shows the collapse is national, not local
to the 2019 earthquake.

\textbf{Conclusions.}
The 2022 collapse is a structural break in the risk-generating process, not a weaker
cohort. We release a calibrated, interpretable screener, framed as associational; a naive
threshold over-flags the poorest quintile (false-positive-rate gap~0.63), which
group-specific thresholds close to~0.003, and partial-identification bounds under 79\%
coverage keep the crisis far above 2018.
\end{minipage}

\vspace{0.7cm}
\begin{minipage}{0.86\linewidth}
\small
\textbf{Keywords:} PISA; educational data mining; low mathematics proficiency; covariate
shift; survey-weighted machine learning; algorithmic fairness; conformal prediction;
Albania
\end{minipage}

\vspace{0.9cm}
{\large July 2026\par}
\end{titlepage}

% =========================== CONTENTS ===========================
\tableofcontents
\thispagestyle{fancy}
\listoffigures
\listoftables

% =========================== BODY ===========================
% Continuous line numbering for the manuscript body (journal requirement).
\linenumbers
\input{sections/introduction}
\input{sections/related}
\input{sections/data}
\input{sections/methods}
\input{sections/results}
\input{sections/context}
\input{sections/discussion}
\input{sections/conclusion}

% =========================== BACK MATTER ===========================
\section*{List of abbreviations}
\addcontentsline{toc}{section}{List of abbreviations}
\nolinenumbers
\begingroup
\small
\begin{tabbing}
\hspace*{2.6cm}\= \kill
ANXMAT \> Mathematics anxiety index\\
AUC \> Area under the receiver operating characteristic curve\\
BELONG \> Sense of belonging (at school) index\\
BRR \> Balanced repeated replication\\
CI \> Confidence interval\\
CV \> Cross-validation\\
DiD \> Difference-in-differences\\
ECE \> Expected calibration error\\
ESCS \> (PISA index of) economic, social and cultural status\\
FPR \> False-positive rate\\
GBM \> Gradient boosting machine\\
GDP \> Gross domestic product\\
HISCED \> Highest parental education (in ISCED levels)\\
HISEI \> Highest parental occupational status (ISEI)\\
HOMEPOS \> Home possessions index\\
ICC \> Intraclass correlation coefficient\\
ICE \> Individual conditional expectation\\
ICT \> Information and communication technology\\
IMMIG \> Immigrant background\\
ISCED \> International Standard Classification of Education\\
ISEI \> International Socio-Economic Index of occupational status\\
KNN \> $k$-nearest neighbours\\
MCC \> Matthews correlation coefficient\\
MLP \> Multilayer perceptron\\
OECD \> Organisation for Economic Co-operation and Development\\
PI \> Prediction interval\\
PISA \> Programme for International Student Assessment\\
PR-AUC \> Area under the precision--recall curve\\
PV \> Plausible value\\
RBF \> Radial basis function\\
ROC \> Receiver operating characteristic\\
SD \> Standard deviation\\
SE \> Standard error\\
SES \> Socioeconomic status\\
SHAP \> SHapley Additive exPlanations\\
SVM \> Support vector machine\\
TPR \> True-positive rate\\
\end{tabbing}
\endgroup
\linenumbers

\section*{Declarations}
\addcontentsline{toc}{section}{Declarations}

\subsection*{Availability of data and materials}
The PISA microdata analysed here are open data, publicly available from the OECD PISA
database (\url{https://www.oecd.org/pisa/data/}). The raw student- and
school-questionnaire files are not redistributed with this work; the full processing,
modelling and analysis pipeline that regenerates every processed slice, table and figure
is available in the project repository at
\url{https://github.com/xhoja/oecd-pisa-albania}.

\subsection*{Competing interests}
The author declares that they have no competing interests.

\subsection*{Funding}
Not applicable.

\subsection*{Authors' contributions}
X.I.\ is the sole author and performed all roles: conceptualisation, data curation,
methodology, software, formal analysis, interpretation, and writing of the manuscript.
The author read and approved the final manuscript.

\subsection*{Acknowledgements}
The author thanks Professor Agresa Qosja for her supervision and guidance throughout this
project, and the Metropolitan University of Tirana for institutional support. This work draws on the
OECD PISA public-use files, and the author thanks the OECD for making these data openly
available.

\subsection*{Authors' information}
Xhoi Ikonomi is an MSc student in Artificial Intelligence (Data Science profile) at the
Metropolitan University of Tirana, Albania.

\bibliographystyle{apalike}
\bibliography{refs}

\end{document}
