\section{Background and Related Work}
\label{sec:related}

\paragraph{PISA survey methodology.}
PISA reports each student's proficiency as ten plausible values drawn from a
posterior latent-trait distribution, and every published statistic is weighted by the
final student weight, with sampling variance estimated by balanced repeated
replication (BRR)~\cite{oecd2024pisa,oecd2009manual}. Rubin's rules combine the point
estimates and the between-imputation variance across the plausible
values~\cite{rubin1987}. These are not optional refinements: using a single averaged
score as a target understates uncertainty, and unweighted statistics do not estimate a
population quantity. A large share of secondary PISA analysis nonetheless drops both
conventions, which, as we argue below, is part of why the accuracy figures it reports
should be read with care. We follow the manual throughout.

\paragraph{Machine learning on PISA background data.}
Predicting achievement or at-risk status from PISA background questionnaires is by now
a well-populated literature, and it converges on a consistent picture in which tree
ensembles dominate. Masci, Johnes and Agasisti~\cite{masci2018student} analyse
PISA~2015 across nine countries with a two-stage tree-based model and recover
socioeconomic status, grade repetition and school climate as leading correlates, while
stressing that much of the variation sits at the school rather than the student level.
On the most recent cycle, Djekourmane et al.~\cite{djekourmane2026east} predict
mathematics scores across six East Asian systems in PISA~2022 and find gradient
boosting strongest: XGBoost reaches $R^2\approx0.57$ and a thresholded AUC near
$0.81$, with LightGBM close behind, and, tellingly, the linear baselines trail badly on
$R^2$ yet come within a few points on AUC. Erdo\u{g}an and
Ta\c{s}tan~\cite{erdogan2024turkish} recover the same ordering on the Turkish PISA~2018
subset, where a boosted regression tree edges out ridge, LASSO, bagging and random
forests. For the post-pandemic window most relevant to us,
Delprato~\cite{delprato2025latam} models low-performing students across ten Latin
American systems in PISA~2022 and uses SHAP attributions to rank their determinants,
foregrounding interpretation over any single accuracy number. Throughout,
XGBoost, LightGBM and CatBoost~\cite{chen2016xgboost,ke2017lightgbm,prokhorenkova2018catboost}
are the recurring strong baselines on this tabular regime, and SHAP
values~\cite{lundberg2017shap} the usual explanation layer.

Two regularities in this work frame our own contribution. First, reported performance
sits in a narrow band, roughly AUC $0.78$ to $0.84$, or accuracy in the high seventies,
largely irrespective of country, cycle, or the exact ensemble chosen. Second, the
strong learners and the linear baselines are usually separated by a fraction of an AUC
point, yet that gap is presented as a ranking rather than tested for significance, and
even studies that expose it directly (linear models trailing on $R^2$ while nearly
tying on AUC~\cite{djekourmane2026east}) stop short of asking whether the difference is
real or what would move it. The literature reports where the accuracy lands but rarely
diagnoses why it lands there.

\paragraph{Accuracy ceilings, and what they are made of.}
That recurring ceiling near AUC~$0.8$ is usually treated as a fixed backdrop, a number
to beat with a better learner or more tuning. We instead make it the object of study,
and this is the paper's central methodological contribution. A plateau in predictive
accuracy can have two very different causes. It may be a \emph{feature ceiling}, in
which a model has exhausted the signal in the feature families it was given but a richer
family (here, the socioeconomic composition of a student's school) would move it; or it
may be a \emph{data ceiling}, in which the measurement instrument itself, the background
questionnaire, cannot resolve risk any further and no feature engineering or model class
will help. The two are indistinguishable from a single learning curve, and the reflex
of tuning harder or adding capacity cannot separate them, because it holds the feature
family fixed. We argue the distinction is nonetheless diagnosable, but only with
\emph{convergent evidence}: a feature ceiling is demonstrated when a new feature family
lifts every model together, and a data ceiling is demonstrated when independent routes
(a hand-crafted composition model, a design-appropriate multilevel model, and a direct
school-questionnaire linkage) all plateau at the same value. We are not aware of a prior
PISA study that draws this line explicitly, which is why the field's shared $\sim0.8$
has gone largely uninterrogated.

\paragraph{Within-country geographic disparities.}
Cross-national PISA work reports one figure per country, and the country-level literature
on Albania documents an urban-rural achievement gap in broadly qualitative
terms~\cite{worldbank_albania_education}: rural students trail urban ones, and the gap is
attributed to socioeconomic development. What is missing, for Albania and for its Balkan
peers, is a \emph{design-based} decomposition of low-proficiency risk \emph{within} the
country. The public-use file carries no subnational region variable, which is likely why
such a decomposition has not been reported; the geographic signal is instead encoded in
the explicit sampling stratum, which labels every school by an urbanicity, region-band and
sector cell. To our knowledge, this paper is the first to recover those cells and report
the survey-weighted below-Level-2 rate for each, with design-based standard errors and a
significance test on every between-group gap, turning a qualitatively known disparity into
a quantified, targetable map. The multilevel literature above establishes that much PISA
risk variance is between schools~\cite{masci2018student,goldstein2011multilevel}; we
locate \emph{which} schools, geographically.

\paragraph{Dataset shift.}
When the input distribution changes between two populations, a classifier trained to
tell them apart quantifies the change, and a high domain-classifier AUC is direct
evidence of covariate shift~\cite{quinonero2009dataset}. We use this as the central
diagnostic for the 2018-to-2022 reversal rather than relying only on marginal score
distributions, which cannot separate a merely lower-scoring cohort from one whose risk
structure has moved.

\paragraph{Calibration, decision curves, and fairness.}
A screening probability must be calibrated to be actionable. Isotonic regression is a
standard non-parametric post-hoc calibrator~\cite{zadrozny2002isotonic}, and
decision-curve analysis evaluates a screener by net benefit across operating thresholds
rather than by accuracy at a single cut~\cite{vickers2006dca}, which matters when
prevalence is high. Group-fairness diagnostics such as demographic parity and equalized
odds expose disparate error rates across protected groups~\cite{hardt2016equalized}. We
report all three, survey-weighted, a combination seldom applied together in the PISA
setting.

\paragraph{Nested design and honest comparison.}
PISA students are nested in schools, so a random-intercept (multilevel) logistic model
is the design-appropriate specification, and its intraclass correlation quantifies the
between-school variance~\cite{goldstein2011multilevel}. Finally, because repeated
cross-validation folds share training data, naive fold-score intervals are
anti-conservative; we use the Nadeau--Bengio corrected-resampled
variance~\cite{nadeau2003inference} and compare models with the corrected resampled
$t$-test (in-sample) and the DeLong test on shared test
predictions~\cite{delong1988}. This is the machinery that turns the field's informal
ranking of near-tied models into a tested claim about which gaps are real.
